### 2022-02-19 introduction/introduction.tex
more

The main contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend these provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve the best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify the internal term representation of E to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and generated less redundant unifiers than the state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 won the higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, the strong basis that we have earlier built made it
  manageable.

  \item After spending so much time and energy on developing higher-order
  provers, we took a step back and worked on improving first-order provers. In
  particular, we looked at the most efficient approaches to simplify problems in
  propositional logic, a simpler, decidable logic that disallows quantification.
  We lifted some of these approaches to first-order logic and implemented them
  in Zipperposition. 
\end{itemize}

@@

The earliest attempts to automate higher-order logic can be traced back to
Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
logic to first-order logic using combinators, similar to the approach taken by
hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
calculi that are directly applied to higher-order formulas.

>>>>

The main contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend these provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve the best possible performance. We discuss those challenges in more detail.

  First, we implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify the internal term representation of E to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

  Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and generated less redundant unifiers than the state of the art.

  We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 won the higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
  prover, Satallax.

  Inspired by this success, we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third phase) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.

  Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, the strong basis that we have earlier built made it
  manageable.

  After spending so much time and energy on developing higher-order
  provers, we took a step back and worked on improving first-order provers. In
  particular, we looked at the most efficient approaches to simplify problems in
  propositional logic, a simpler, decidable logic that disallows quantification.
  We lifted some of these approaches to first-order logic and implemented them
  in Zipperposition.

@@

The earliest attempts to automate higher-order logic can be traced back to
Robinson, who provided two approaches. The first one \cite{ar-69-hol} operates
directly on higher-order formulas. The second approach \cite{ar-70-hol} is based
on translating higher-order logic to first-order logic using combinators,
similar to the approach taken by hammers. Huet~\cite{gh-73-hol} and
Andrews~\cite{pa-71-type-theory} designed calculi that are directly applied to
higher-order formulas.

### 2022-02-19 pre/pre.tex
more

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We finish with the
    description of the superposition calculus. The text of this chapter is
    partially based on the preliminaries sections of the publications listed in
    Chapter \ref{ch:intro}.

@@

\emph{Terms} are the basic building blocks of first-order logic. Variables
$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar
notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
or simply $\tuplen{t}$.

Terms are used to build \emph{atoms}. Atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for any value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic, it does not suffice to assign
values to variables to determine the truth value of the formula. It is also
necessary to interpret symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

@@

{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is {\em grounding}
if it maps all variables to ground terms.

@@

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
\tau_2$, and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
$\tau_2$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau_1
\rightarrow \tau_2$. The syntactic distinction between free and bound variables
gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
\cite{tn-93-patterns}. Out of terms, atoms and formulas are built in the same
way as in first-order logic.

@@

The resolution and superposition calculi do not work directly on formulas, but
on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed in a set of clauses. For all three
discussed logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}.

@@

In standard, non-clausal, first-order and higher-order logic, an atom is either
predicate or equational. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

@@

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
unifiers to higher-order logic, the concept of {\em complete sets of unifiers} was
introduced.

A (higher-order) unifier of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
\in E$. A complete set of unifiers (\emph{CSU}) of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general unifier
corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist algorithms
that enumerate all elements of CSU for unifiable terms.

@@

Before we spell out the inference rules of standard, first-order superposition
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \leftarrow
t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within literals in
\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

\[
\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
\]
where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
for resolution; \infname{ER} stands for equality resolution

\[
\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
\]
where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
$(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring

\[
\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
$(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
\infname{SP} stands for positive superposition

\[
\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
$(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
\infname{SN} stands for negative superposition
\medskip

@@

Simple rules such as removing duplicate literals, literals of the form $s \noteq
s$ or tautological clauses are clearly justified by the redundancy criterion.
Schulz gives an extensive list of such rules \cite{ss-02-brainiac}.

@@

Superposition provers saturate the input problem with respect to the calculus's
inference rules using the \emph{given clause procedure}
\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
passive set $\mathcal{P}$ and an active set $\mathcal{A}$. All clauses start in
$\mathcal{P}$. At each iteration of the procedure's main loop, the prover
chooses a clause $C$ from $\mathcal{P}$, simplifies it, and moves it to
$\mathcal{A}$. Then all inferences between $C$ and active clauses are performed.
The resulting clauses are again simplified and put in $\mathcal{P}$.
The provers differ in which clauses are used for simplification: Otter-loop
\cite{mcw-1997-otter} provers use both active and passive clauses whereas
DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.
The provers we discuss in this thesis are both DISCOUNT-loop provers.

>>>>

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We finish with the
    description of the superposition calculus. As this thesis discusses practical
    aspects of theorem proving, we define only the fundamental notions, while
    more advanced notions are intuitively described with references to rigorous
    definitions. The text of this chapter is
    partially based on the preliminaries sections of the publications listed in
    Chapter \ref{ch:intro}.

@@

\emph{Terms} are the basic building blocks of first-order logic, built inductively as follows. Variables
$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
of type $\tau$. If $n=0$, we drop the parentheses and write $\cst{f}$. We also
abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using a similar
notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
or simply $\tuplen{t}$.

Terms are used to build \emph{atoms}. Atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for any value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic, it does not suffice to assign
values to variables to determine the truth value of the formula. It is also
necessary to interpret the symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

@@

{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution $\sigma$ is {\em grounding}
if $\sigma(x) \not= x$ implies that $\sigma(x)$ is ground.

@@

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
\tau_2$ and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
$\tau_2$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau_1
\rightarrow \tau_2$. The syntactic distinction between free and bound variables
gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
\cite{tn-93-patterns}. Note that there is no requirement that a symbol is not
applied to terms of Boolean type as in first-order logic. Furthermore, all
logical connectives are symbols present in the set $\Sigma$. This means that
higher-order logic does not distinguish between terms and formulas. For example,
$\cst{p} \iand \cst{q}$ and $\cst{f} \, (\cst{p} \iand \cst{q})$ are well-formed
terms of higher-order logic. For convenience, we still call terms of Boolean
types \emph{formulas}.

@@

The resolution and superposition calculi do not work directly on formulas, but
on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed in a set of clauses. For all three
discussed logics, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}.

@@

In standard, non-clausal first-order logic, an atom is either
predicate or equational. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

@@

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
is more general than the other. To generalize most general
unifiers to higher-order logic, the concept of tje {\em complete sets of unifiers} (CSU) was
introduced.

A (higher-order) unifier of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma(s) \equi \sigma(t)$, for all $s \unif t
\in E$. A complete set of unifiers of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho(X) = \theta(\sigma(X))$. The most general unifier
corresponds to a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist algorithms
that enumerate all elements of CSU for unifiable terms.

@@

Before we spell out the inference rules of standard, first-order superposition
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \leftarrow
t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within the literals in
\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality resolution} (ER)} \\[\jot]
    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible for resolution
\end{tabular}

\medskip

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality factoring} (EF)} \\[\jot]
    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}$ 
        & where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
        $(s \eq t)\sigma$ is eligible for paramodulation
\end{tabular}

\medskip

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into positive literals} (SP)} \\[\jot]
    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
        $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable
\end{tabular}

\medskip

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into negative literals} (SN)} \\[\jot]
    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
         $(s \eq t)\sigma$ is eligible for paramodulation, $(u \noteq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable
\end{tabular}

% \[
% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
% \]
% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
% for resolution; \infname{ER} stands for equality resolution

% \[
% \namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
% \]
% where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
% for resolution; \infname{ER} stands for equality resolution

% \[
% \namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
% \]
% where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
% $(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring

% \[
% \namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
% \]
% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
% $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
% \infname{SP} stands for positive superposition

% \[
% \namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}
% \]
% where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
% $(s \eq t)\sigma$ is eligible for paramodulation, $(u \eq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable;
% \infname{SN} stands for negative superposition
\medskip

@@

Simple rules such as removing duplicate literals, literals of the form $s \noteq
s$ or tautological clauses are clearly justified by the redundancy criterion.
Schulz gives an extensive list of such rules \cite{ss-02-brainiac}, including rewriting of positive
literals which has more side conditions.

@@

Superposition provers saturate the input problem with respect to the calculus's
inference rules using the \emph{given clause procedure}
\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
passive set $\mathcal{P}$ and an active set $\mathcal{A}$. All clauses start in
$\mathcal{P}$. At each iteration of the procedure's main loop, the prover
chooses a clause $C$ from $\mathcal{P}$, simplifies it, and moves it to
$\mathcal{A}$. Then all inferences between $C$ and active clauses are performed.
The resulting clauses are again simplified and put in $\mathcal{P}$.
The provers differ in which clauses are used for simplification: Otter-loop
\cite{mcw-1997-otter} provers use both active and passive clauses whereas
DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.
The provers we discuss in this thesis, E and Zipperposition, are both DISCOUNT-loop provers.
